---
title: "amerifluxr demo"
author: "Housen Chu"
output: rmarkdown::html_document
---

```{r setup, warning = FALSE}
library(amerifluxr)
library(data.table)
library(pander)
```

amerifluxr is a programmatic interface to [AmeriFlux](https://ameriflux.lbl.gov/). 
This vignette demonstrates examples to query a list of target sites based on 
sites' general information and availability of metadata and data, and then 
download, import, and parse the data and metadata.     

## Get a site list with general info

AmeriFlux data are organized by individual sites. Typically, data query begins 
with site search and selection. A full list of AmeriFlux sites with general info 
can be obtained using the amf_site_info() function. 

Convert the site list to a data.table for easier manipulation. Also see 
[link](https://ameriflux.lbl.gov/data/badm/badm-standards/) for variable 
definition. 

```{r echo=TRUE, results = "asis"}
# get a full list of sites with general info
sites <- amf_site_info()
sites_dt <- data.table::as.data.table(sites)

#pander::pandoc.table(sites_dt[c(1:3), ])

```

The site list provides a summary of all registered sites and sites with data.

It's often important to understand the data use policy under which the data are 
shared. In 2021, the AmeriFlux community moved to CC-BY-4.0 License. Most site 
PIs now share their data under the CC-BY-4.0 license. Data for some sites are 
shared under the historical AmeriFlux data-sharing policy, now called the 
AmeriFlux Legacy Data Policy. 

Check [link](https://ameriflux.lbl.gov/data/data-policy/#data-use) for data use 
policy and attribution guidelines.

```{r results = "asis"}
# total number of registered sites
pander::pandoc.table(sites_dt[, .N])

# total number of sites with available data
pander::pandoc.table(sites_dt[!is.na(DATA_START), .N])

# get number of sites with available data, grouped by data use policy
pander::pandoc.table(sites_dt[!is.na(DATA_START), .N, by = .(DATA_POLICY)])

```

Once decided, we can query a target site list based on the desired criteria, 
e.g., IGBP, data availability, data policy. 

```{r results = "asis"}
# get a list of cropland and grassland sites with available data,
#  shared under CC-BY-4.0 data policy, data available in 2017-
crop_ls <- sites_dt[IGBP %in% c("CRO", "GRA", "CVM") &
                      !is.na(DATA_START) & DATA_START <= 2021 & DATA_END >= 2017 &
                      DATA_POLICY == "CCBY4.0", ]

#pander::pandoc.table(crop_ls[c(1:10),])

```

## Get metadata availability

Next, we want to know if certain types of metadata are available for the 
selected sites. The amf_list_metadata() function provides a summary of metadata 
availability before actually downloading the data and metadata. It returns the 
number of entries of each BADM variable/group for each site. The site_set 
parameter of amf_list_data() can be used to subset the sites of interest. 

Check [AmeriFlux webpage](https://ameriflux.lbl.gov/data/badm/badm-basics/)
for definitions of all BADM groups.     

```{r results = "asis"}
# retrieve the metadata availability for selected sites
metadata_aval_sub <- as.data.table(amf_list_metadata(site_set = crop_ls$SITE_ID))

# down-select sites by availability of canopy height (GRP_HEIGHTC)
crop_ls2 <- metadata_aval_sub[GRP_HEIGHTC > 0, .(SITE_ID, GRP_HEIGHTC)][order(-GRP_HEIGHTC)]

pander::pandoc.table(crop_ls2)
```

## Get data availability

Next, we can use amf_list_data() to query the availability of specific variables
in the data (i.e., flux/met data, so-called BASE data product).

The amf_list_data() provides a summary of variable availability (per site/year)
before downloading the data. The site_set parameter of amf_list_data() can be 
used to subset the sites of interest. 

```{r results = "asis"}
# get data availability for selected sites
data_aval <- data.table::as.data.table(amf_list_data(site_set = crop_ls2$SITE_ID))

#pander::pandoc.table(data_aval[c(1:10), ])
```

We can use the variable availability to subset sites that have certain variables
in specific years. 

The BASENAME column indicates the variable's base name (i.e., ignoring position 
qualifier), and can be used to get a coarse-level variable availability. 

See [AmeriFlux website](https://ameriflux.lbl.gov/data/aboutdata/data-variables/)
for definitions of base names and qualifiers.

```{r results = "asis"}
# down-select sites based on the available target variables in 2017-2021
data_aval_sub <- data_aval[data_aval$BASENAME %in% c("LE", "TA", "RH", "LW_OUT") &
                             !data_aval$GAP_FILLED,
                           .(SITE_ID, VARIABLE, BASENAME, 
                             Y2017, Y2018, Y2019, Y2020, Y2021,
                             Y2017_2021 = (Y2017 + Y2018 + Y2019 + Y2020 + Y2021) / 5)]

crop_ls3 <- data_aval_sub[data_aval_sub$Y2017_2021 > 0, ]

#pander::pandoc.table(table(crop_ls3$SITE_ID, crop_ls3$BASENAME))

```

A companion function amf_plot_datayear() can be used for visualizing the data 
availability in an interactive figure. However, it is strongly advised to subset
the sites, variables, and/or years for faster processing and better visualization.

```{r}
# plot data availability of LE for selected sites in 2017-2021
#  use unique() to remove redundant SITE_ID in cropl_ls3 
amf_plot_datayear(
  site_set = unique(crop_ls3$SITE_ID),
  var_set = c("LE"),
  nonfilled_only = TRUE,
  year_set = c(2017:2021)
)

```

## Download data

AmeriFlux data and metadata can be downloaded using amf_download_base() and 
amf_download_bif(). Users will need to create a personal AmeriFlux account 
[here](https://ameriflux-data.lbl.gov/Pages/RequestAccount.aspx) before download. 

The following downloads AmeriFlux flux/met data (aka BASE data product) from 
multiple sites.

```{r eval = FALSE}
## finalize the list of sites to download
final_list <- unique(crop_ls3$SITE_ID)
final_list <- final_list[-which(final_list == "US-Twt")]
  
## When running, replace user_id and user_email with a real AmeriFlux account
floc2 <- amf_download_base(
  user_id = Sys.getenv("MY_USER"),  ## replace with your user ID
  user_email = Sys.getenv("MY_EMAIL"), ## replace with your email
  site_id = final_list,
  data_product = "BASE-BADM",
  data_policy = "CCBY4.0",
  agree_policy = TRUE,
  intended_use = "other",
  intended_use_text = "amerifluxr package demonstration",
  verbose = TRUE,
  out_dir = tempdir()
)

print(floc2)
```

The downloaded files are zipped files by individual sites saved in tempdir()
(e.g., AMF\_\{SITE_ID\}\_BASE-BADM\_\{VERSION\}\.zip), which contains a BASE 
data file and a metadata file (aka BADM data product. The amf_download_base() 
also returns the file path to the downloaded file, which can be used later to 
read the file into R.

# BASE data product
## Import data

The amd_read_base() imports a BASE file, either from a zipped file
or an unzipped comma-separated file (.csv). The parse_timestamp 
parameter can be used if additional time-keeping columns (e.g., 
year, month, day, hour) are desired. 

```{r results = "asis"}
# read the BASE from a zip file, without additional parsed time-keeping columns
base1 <- amf_read_base(
  file = floc2[1],
  unzip = TRUE,
  parse_timestamp = TRUE
)

plot(base1$TIMESTAMP, 
     base1$LE_1_1_1,
     pch = 16,
     cex = 0.5,
     col = rgb(0, 0, 0, 0.1))

```

## Parse and interpret data

The details of the BASE data product's format and variable definitions are on 
[AmeriFlux website](https://ameriflux.lbl.gov/data/aboutdata/data-variables/).

In short, the BASE data product contains flux, meteorological, and soil 
observations that are reported at regular intervals of time, generally 
half-hourly or hourly, for a certain time period. **TIMESTAMP_START**
and **TIMESTAMP_END** columns (i.e., YYYYMMDDHHMM 12 digits) denote the starting 
and ending time of each reporting interval (i.e., row). 

All other variables use the format of \{base name\}\_\{qualifier\}, e.g., FC_1, 
CO2_1_1_1. Base names indicate fundamental quantities that are either measured 
or calculated / derived. Qualifiers are suffixes appended to variable base names
that provide additional information (e.g., gap-filling, position) about the 
variable. In some cases, qualifiers are omitted if only one variable is provided
for a site.

The amf_variable() retrieves the latest list of base names and default units. 
For sites that have relatively fewer variables and less complicated qualifiers, 
the users could easily interpret variables and qualifiers. 

```{r results = "asis"}
# get a list of latest base names and units. 
FP_ls <- amf_variables()

#pander::pandoc.table(FP_ls[c(11:20), ])
```

Alternatively, the amf_parse_basename() can programmatically parse the the 
variable names into base names and qualifiers.

This function can be helpful for sites with many variables and relatively 
complicated qualifiers, as a prerequisite for handling data from many sites. 
The function returns a data frame with information about each variable's base 
name, qualifier, and whether a variable is gap-filled, layer-aggregated, or 
replicate aggregated.

```{r results = "asis"}
# parse the variable name
basename_decode <- amf_parse_basename(var_name = colnames(base1))

#pander::pandoc.table(basename_decode[c(1, 2, 3, 4, 6, 11, 12),])
```

## Data filtering

While BASE data products are quality-checked before release, the data may not be 
filtered for all outliers. The amf_filter_base() can be use to filter the data 
based on the expected physically ranges (i.e., obtained through amf_variables()). 

By default, a ±5% buffer is applied to account for possible edge values near 
the lower and upper bounds, which are commonly observed for certain variables
like radiation, relative humidity, and snow depth. 

```{r }
# filter data, using default physical range +/- 5% buffer
base_f <- amf_filter_base(data_in = base1)

```

## Measurement height information

Measurement height information contains height/depth and instrument model 
information of the BASE data products. The info can be downloaded directly using
the amf_var_info() function. The function returns a data frame for all available 
sites, and can be subset using the "Site_ID" column. The "Height" column refers
to the distance from the ground surface in meters. Positive values are heights,
and negative values are depths. 

See [web page](https://ameriflux.lbl.gov/data/measurement-height/) for details. 

```{r results = "asis"}
# obtain the latest measurement height information
var_info <- amf_var_info()

# subset the variable by target Site ID
var_info <- var_info[var_info$Site_ID %in% final_list, ]
#pander::pandoc.table(var_info[c(1:10), ])
```

# BADM data product
## Download and Import BADM data

Biological, Ancillary, Disturbance, and Metadata (BADM) are non-continuous 
information that describe and complement continuous flux and meteorological data 
(e.g., BASE data product). BADM include general site description, metadata about
the sensors and their setup, maintenance and disturbance events, and biological 
and ecological data that characterize a site. 

See [link](https://ameriflux.lbl.gov/data/badm/badm-basics/) for details.

The following downloads a single file containing all AmeriFlux sites' metadata 
(i.e., BADM data product) for sites under the CC-BY-4.0 data use policy. 

The downloaded file is a Excel file saved to tempdir(). Similarly, the 
amf_download_bif() also returns the file path to the downloaded file, which can 
be used later to read the file into R.

```{r eval = FALSE}
## When running, replace user_id and user_email with a real AmeriFlux account
floc1 <- amf_download_bif(
  user_id = Sys.getenv("MY_USER"), ## replace with your user ID
  user_email = Sys.getenv("MY_EMAIL"), ## replace with your email
  data_policy = "CCBY4.0",
  agree_policy = TRUE,
  intended_use = "other",
  intended_use_text = "amerifluxr package demonstration",
  out_dir = tempdir(),
  verbose = TRUE,
  site_w_data = TRUE
)

print(floc1)
```

The amf_read_bif() can be used to import the BADM data file. It returns a data 
frame for all available sites, and can subset using the "SITE_ID" column. 

```{r results = "asis"}
# read the BADM BIF file, using an example data file
bif <- amf_read_bif(file = floc1)

# subset by target Site ID
bif <- bif[bif$SITE_ID %in% final_list, ]
#pander::pandoc.table(bif[c(1:15), ])

# get a list of all BADM variable groups and variables
unique(bif$VARIABLE_GROUP)
length(unique(bif$VARIABLE))

```

As shown above, BADM data contain information from a variety of variable groups 
(i.e., GRP\_\{BADM_GROUPS\}). Browse the definitions of all available variable 
groups [here](https://ameriflux.lbl.gov/data/badm/badm-standards/).  

To get the BADM data for a certain variable group, use amf_extract_badm(). 
The function also renders the data format (i.e., display variables by columns)
for human readability.

```{r results = "asis"}
# extract the HEIGHTC (canopy height) group
bif_hc <- amf_extract_badm(bif_data = bif, select_group = "GRP_HEIGHTC")
#pander::pandoc.table(bif_hc)
```

Note: amf_extract_badm() returns all columns in characters. Certain groups of 
BADM variables contain columns of timestamps (i.e., ISO format) and data values, 
and need to be converted before further use.     

```{r fig.width = 7}
# convert HEIGHTC_DATE to POSIXlt
bif_hc$TIMESTAMP <- strptime(bif_hc$HEIGHTC_DATE, format = "%Y%m%d", tz = "GMT")

# convert HEIGHTC column to numeric
bif_hc$HEIGHTC <- as.numeric(bif_hc$HEIGHTC)

# plot time series of canopy height
plot(bif_hc$TIMESTAMP[bif_hc$SITE_ID == "US-ARM"],
     bif_hc$HEIGHTC[bif_hc$SITE_ID == "US-ARM"], 
     xlab = "TIMESTAMP",
     ylab = "canopy height (m)")

```

Last, the contacts of the site members and data DOI can be obtained from the 
BADM. The AmeriFlux [data policy](https://ameriflux.lbl.gov/data/data-policy/) 
requires proper attribution (e.g., data DOI). 

In some case, for example, using data shared under Legacy Data Policy for 
publication, data users are required to contact data contributors directly, so 
that they have the opportunity to contribute substantively and become a co-author. 

```{r results = "asis"}
# get a list of contacts
bif_contact <- amf_extract_badm(bif_data = bif, select_group = "GRP_TEAM_MEMBER")
#pander::pandoc.table(bif_contact[c(1:3), ])

# get data DOI
bif_doi <- amf_extract_badm(bif_data = bif, select_group = "GRP_DOI")
#pander::pandoc.table(bif_doi[c(1:3), ])
```

